$p$-adic precision

### 2014-02-07 padicprec.tex

If we denote by $NP_F$ the Newton polygon application of the power serie $F(t) = \sum_{k\geq 0}^{+\infty} \interleave L_{k+1} \interleave  t^k$, then it is enough that :
\[ \sup_{ x \in \mathbb{R}_+} \left( NP_F(x) -x  \log_p l \right) \geq - \log_p \varepsilon.\]

@@

Suppose that we are given a polynomial $P_0 \in K_d[X]$ written as a 
product $P_0 = A_0 B_0$ where $A_0$ and $B_0$ are monic and coprime. 
Hensel's lemma implies that there exists a small neighborhood $\mathcal 
U$ of $P_0$ in $K_d[X]$ such that any $P \in \mathcal U$ factors 
uniquely as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ 
and $B_0$ respectively. Thus, we can consider the application $f : P 
\mapsto (A,B)$ defined on the subset of $\mathcal U$ consisting of monic 
polynomials. We want to differentiate $f$ at $P_0$. For this, we 
differentiate the equality $P = A B$ around $P_0$, obtaining
\begin{equation}
\label{eq:difffactor}
dP = A_0 \cdot dB + B_0 \cdot dA.
\end{equation}
where $dP$, $dA$ and $dB$ have degree less than $\deg P$, $\deg A$ and 
$\deg B$ respectively. If $A_0 U_0 + B_0 V_0 = 1$ is a Bezout relation
between $A_0$ and $B_0$, it follows from Eq.~\eqref{eq:difffactor} that
$dA$ (resp. $dB$) is the remainder in the Euclidean division of $V_0
{\cdot} dP$ by $A_0$ (resp. of $U_0 {\cdot} dP$ by $B_0$).

@@

An important special case of the previous study occurs when $A_0$ has 
degree $1$, that is $A_0(X) = X - \alpha_0$ with some $\alpha_0 \in K$. 
The application $P \mapsto A$ is then nothing but the mapping that 
follows the simple root $\alpha_0$. Of course, its differential around 
$P_0$ can be computed by the above method. Nevertheless there is a more 
direct way to obtain it in this particular case: we write $(P_0 + 
dP)(\alpha_0 + d\alpha) = 0$ and expand this. We obtain this way 
$P_0'(\alpha_0) d\alpha + dP(\alpha_0) = 0$ where $P'$ denotes the 
derivative of $P$. Since $\alpha_0$ is a simple root, $P_0'(\alpha_0)$ 
does not vanish and we find:
$$d \alpha = - \frac{dP(\alpha_0)}{P'_0(\alpha_0)}.$$

@@

We then have a notion of division in $K[\XX]$: if $f, f_1, \ldots, f_s$ 
are polynomials in $K[\XX]$, there exists a writing:
$$f = q_1 f_1 + \cdots + q_s f_s + r$$
where $q_1, \ldots, q_s, r \in K[\XX]$ and no term of $r$ is divisible
by the leading term of some $f_i$ ($1 \leq i \leq s$). Moreover, assuming
that $(f_1, \ldots, f_s)$ is a Gr\"obner basis\footnote{Without this
assumption, there still exists a canonical choice of $r$. We do not know
if the computation of the differential that follows extends to a more
general setting.} (of the ideal generated by these polynomials), the 
polynomial $r$ is uniquely determined and called the \emph{remainder} of 
the division of $f$ by the family $(f_1, \ldots, f_s)$. The application 
$(f, f_1, \ldots, f_s) \mapsto r$ is then well defined and similarly to 
the case of univariate polynomials, we can compute its differential: we 
find that $dr$ is obtained as the remainder of the division of $f - (q_1 
\cdot d f_1 + \cdots + q_s \cdot d f_s)$ by $(f_1, \ldots, f_s)$.

@@

Recall that any 
ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We can 
then consider the application mapping a family $(f_1, \ldots, f_s)$ of 
\emph{homogeneous}\footnote{This restriction will be convenient for us 
mainly because we are using the article \cite{Vaccon}, which deals only 
with homogeneous polynomials. It is however probably not essential.} 
polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
\ldots, g_t)$. It follows from Theorem 1.1 of \cite{Vaccon} that this 
application at $(f_1, \ldots, f_s)$ is continuous provided that:
\begin{itemize}
\item the sequence $(f_1, \ldots, f_s)$ is regular
\item for all $i \in \{1, \ldots, s\}$, the ideal generated by
$f_1, \ldots, f_i$ is wearly-$w$ where $w$ denotes the fixed monomial
order (\emph{cf} \cite{Vaccon} for more precisions).
\end{itemize}
A similar argument proves that it is actually differentiable at such
points. We are now going to compute the differential. For this we
remark that since the families $(f_1, \ldots, f_s)$ and $(g_1, \ldots,
g_t)$ generate the same ideal, we have a relation:
\begin{equation}
\label{eq:prodgrobner}
(g_1, \ldots, g_t) = (f_1, \ldots, f_s) \cdot A
\end{equation}
where $A$ is a $(s \times t)$ matrix with coefficients in $K[\XX]$.
Moreover following Vaccon's construction, we see that the entries of $A$ 
can be chosen in such a way that they define differentiable functions.
Differentiating Eq.~\eqref{eq:prodgrobner}, we get:
$$(d g_1, \ldots, d g_t) =
(d f_1, \ldots, d f_s) \cdot A + (f_1, \ldots, f_s) \cdot dA.$$
Finally, using that we are working with \emph{reduced} Gr\"obner
basis, the above equality implies that $d g_i$ is the remainder in
the division of the $i$-th coefficient of the product
$(d f_1, \ldots, d f_s) \cdot A$ by the family $(g_1, \ldots, g_t)$.

@@

We consider the open subset $\mathcal V_{n-d}$ of $M_{n,n-d}(K)$ 
consisting of matrices of full rank, \emph{i.e.} rank $n-d$. The left 
kernel defines a mapping $\text{LK} : M_{n,n-d}(K) \to \Grass(d,n)$. Let 
us prove that it is differentiable and compute its differential around 
some point $M \in M_{n,n-d}(K)$. Of course, there exists a neighborhood 
$\mathcal U$ of $M$ whose image is entirely contained in a given chart 
$U_P$. We assume for simplicity that $P$ is the identity. On $\mathcal 
U$, the map $\text{LK}$ corresponds in our chart to the application that 
takes $M$ to the unique matrix $N \in M_{d, n-d}(K)$ such that 
$\begin{pmatrix} I_d & N \end{pmatrix} \cdot M = 0$. The implicit 
function theorem then implies that $\text{LK}$ is differentiable. 
Furthermore, its differential satisfies the relation
$\begin{pmatrix} 0 & dN \end{pmatrix} \cdot M +
\begin{pmatrix} I_d & N \end{pmatrix} \cdot dM = 0$,
from which we can compute $dN$ by projecting on the $(n-d)$ last
columns.

>>>>

If we denote by $NP_F$ the Newton polygon of the power series 
\[
F(t) = \sum_{k\geq 0}^{+\infty} \interleave L_{k+1} \interleave  t^k
\]
then it is enough that:
\[
\sup_{ x \in \mathbb{R}_+} \left( NP_F(x) -x  \log_p l \right) \geq - \log_p \varepsilon.
\]

@@

Suppose that we are given a polynomial $P_0 \in K_d[X]$ written as a 
product $P_0 = A_0 B_0$ where $A_0$ and $B_0$ are monic and coprime. 
Hensel's lemma implies that there exists a small neighborhood $\mathcal 
U$ of $P_0$ in $K_d[X]$ such that any $P \in \mathcal U$ factors 
uniquely as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ 
and $B_0$ respectively. Thus, we can consider the map $f : P 
\mapsto (A,B)$ defined on the subset of $\mathcal U$ consisting of monic 
polynomials. We want to differentiate $f$ at $P_0$. For this, we 
differentiate the equality $P = A B$ around $P_0$, obtaining
\begin{equation}
\label{eq:difffactor}
dP = A_0 \cdot dB + B_0 \cdot dA.
\end{equation}
where $dP$, $dA$ and $dB$ have degree less than $\deg P$, $\deg A$ and 
$\deg B$ respectively. If $A_0 U_0 + B_0 V_0 = 1$ is a Bezout relation
between $A_0$ and $B_0$, it follows from Eq.~\eqref{eq:difffactor} that
$dA$ (resp. $dB$) is the remainder in the Euclidean division of $V_0
{\cdot} dP$ by $A_0$ (resp. of $U_0 {\cdot} dP$ by $B_0$).

@@

An important special case of the previous study occurs when $A_0$ has 
degree $1$, that is $A_0(X) = X - \alpha_0$ with some $\alpha_0 \in K$. 
The map $P \mapsto A$ is then nothing but the mapping that 
follows the simple root $\alpha_0$. Of course, its differential around 
$P_0$ can be computed by the above method. Nevertheless there is a more 
direct way to obtain it in this particular case: we write $(P_0 + 
dP)(\alpha_0 + d\alpha) = 0$ and expand this. We obtain this way 
$P_0'(\alpha_0) d\alpha + dP(\alpha_0) = 0$ where $P'$ denotes the 
derivative of $P$. Since $\alpha_0$ is a simple root, $P_0'(\alpha_0)$ 
does not vanish and we find:
$$d \alpha = - \frac{dP(\alpha_0)}{P'_0(\alpha_0)}.$$

@@

We then have a notion of division in $K[\XX]$: if $f, f_1, \ldots, f_s$ 
are polynomials in $K[\XX]$, there exists a writing:
$$f = q_1 f_1 + \cdots + q_s f_s + r$$
where $q_1, \ldots, q_s, r \in K[\XX]$ and no term of $r$ is divisible
by the leading term of some $f_i$ ($1 \leq i \leq s$). Moreover, assuming
that $(f_1, \ldots, f_s)$ is a Gr\"obner basis\footnote{Without this
assumption, there still exists a canonical choice of $r$. We do not know
if the computation of the differential that follows extends to a more
general setting.} (of the ideal generated by these polynomials), the 
polynomial $r$ is uniquely determined and called the \emph{remainder} of 
the division of $f$ by the family $(f_1, \ldots, f_s)$. The map 
$(f, f_1, \ldots, f_s) \mapsto r$ is then well defined and similarly to 
the case of univariate polynomials, we can compute its differential: we 
find that $dr$ is obtained as the remainder of the division of $f - (q_1 
\cdot d f_1 + \cdots + q_s \cdot d f_s)$ by $(f_1, \ldots, f_s)$.

@@

Recall that any 
ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We can 
then consider the map sending a family $(f_1, \ldots, f_s)$ of 
\emph{homogeneous}\footnote{This restriction will be convenient for us 
mainly because we are using the article \cite{Vaccon}, which deals only 
with homogeneous polynomials. It is however probably not essential.} 
polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
\ldots, g_t)$. It follows from Theorem 1.1 of \cite{Vaccon} that this 
map at $(f_1, \ldots, f_s)$ is continuous provided that:
\begin{itemize}
\item the sequence $(f_1, \ldots, f_s)$ is regular
\item for all $i \in \{1, \ldots, s\}$, the ideal generated by
$f_1, \ldots, f_i$ is wearly-$w$ where $w$ denotes the fixed monomial
order (\emph{cf} \cite{Vaccon} for more precisions).
\end{itemize}
A similar argument proves that it is actually differentiable at such
points. We are now going to compute the differential. For this we
remark that since the families $(f_1, \ldots, f_s)$ and $(g_1, \ldots,
g_t)$ generate the same ideal, we have a relation:
\begin{equation}
\label{eq:prodgrobner}
(g_1, \ldots, g_t) = (f_1, \ldots, f_s) \cdot A
\end{equation}
where $A$ is a $(s \times t)$ matrix with coefficients in $K[\XX]$.
Moreover following Vaccon's construction, we see that the entries of $A$ 
can be chosen in such a way that they define differentiable functions.
Differentiating Eq.~\eqref{eq:prodgrobner}, we get:
$$(d g_1, \ldots, d g_t) =
(d f_1, \ldots, d f_s) \cdot A + (f_1, \ldots, f_s) \cdot dA.$$
Finally, using that we are working with \emph{reduced} Gr\"obner
basis, the above equality implies that $d g_i$ is the remainder in
the division of the $i$-th coefficient of the product
$(d f_1, \ldots, d f_s) \cdot A$ by the family $(g_1, \ldots, g_t)$.

@@

We consider the open subset $\mathcal V_{n-d}$ of $M_{n,n-d}(K)$ 
consisting of matrices of full rank, \emph{i.e.} rank $n-d$. The left 
kernel defines a mapping $\text{LK} : M_{n,n-d}(K) \to \Grass(d,n)$. Let 
us prove that it is differentiable and compute its differential around 
some point $M \in M_{n,n-d}(K)$. Of course, there exists a neighborhood 
$\mathcal U$ of $M$ whose image is entirely contained in a given chart 
$U_P$. We assume for simplicity that $P$ is the identity. On $\mathcal 
U$, the map $\text{LK}$ corresponds in our chart to the map that 
takes $M$ to the unique matrix $N \in M_{d, n-d}(K)$ such that 
$\begin{pmatrix} I_d & N \end{pmatrix} \cdot M = 0$. The implicit 
function theorem then implies that $\text{LK}$ is differentiable. 
Furthermore, its differential satisfies the relation
$\begin{pmatrix} 0 & dN \end{pmatrix} \cdot M +
\begin{pmatrix} I_d & N \end{pmatrix} \cdot dM = 0$,
from which we can compute $dN$ by projecting on the $(n-d)$ last
columns.
